---
title: "Custom types"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(torchexport)
```

When extending torch with C++, you might want your C++ function to return types that are not defined in [torch.h](https://github.com/mlverse/torch/blob/93b250ae48911f572d5477e8f52b19c40a5da86c/inst/include/torch_types.h#L659-L763), or ones where no automatic casting from and to `raw` pointers has yet been implemented in [lantern/types.h](https://github.com/mlverse/torch/blob/93b250ae48911f572d5477e8f52b19c40a5da86c/lantern/include/lantern/types.h#L50-L113). `torchexport` can be extended to support custom types defined in user packages, provided that you implement all the necessary casting:

-   `frow_raw` and `make_raw` : On the `csrc` side, you must specify how your type is to be converted to a `void*` pointer, as well as how to convert from `void*` to your custom type.
-   `Rcpp` type: On the Rcpp side ,you need to implement the casting between `void*` and a custom Rcpp type that can manage the memory pointed by this `void*`.
-   Rcpp type to `SEXP`: if you ever return this type to R (as opposed to only using it in Rcpp) you need to implement casting from the Rcpp custom type to a `SEXP`.
-   `SEXP` to Rcpp type: if you use this type as an argument on the R side, you must implement the casting from `SEXP` to your custom type.

## Example

In this example we will implement the `tensor_pair` type, an alias for `std::tuple<torch::Tensor, torch::Tensor>`. This type is not implemented in `torch`, thus we need to implement it ourselves. In fact, this type was needed in the [torchsparse](https://github.com/mlverse/torchsparse) package; that's why the names of the files will sometimes start with `torchsparse_`.

### The `csrc` side

First we will create a file called `torchparse_types.h` in the `csrc/include/torchsparse` directory. This file will contain the declaration of your custom type and the declarations for the functions that allows casting this type to and from `void*` pointers.

For example, it can look like this:

``` cpp
#include <torch/torch.h>

// declares the alias, but could also be a `class CustomType`.
using tensor_pair = std::tuple<torch::Tensor,torch::Tensor>;

// In this namespace we declare the function that creates a `void*`
// from an instance of your type. This `void*` pointer must own all
// its memory.
namespace make_raw {
  void* TensorPair (const tensor_pair& x);
}

// In this namespace we declare a function that takes a void* pointer and
// returns a reference to your type. It's a good idea to return by 
// reference.
namespace from_raw {
  tensor_pair& TensorPair (void* x);
}
```

Now that we declared the type and it's casting functions, we will implement these functions. The name of the file doesn't matter here, but it's nice to use `torchsparse_types.cpp`. This file lives in `csrc/src`. Don't forget to add it to the `CMakeLists.txt` file so it also gets compiled.

The implementation of `make_raw::TensorPair` and `from_raw::TensorPair` look like this:

``` cpp
#include <torch/torch.h>
#include "torchsparse/torchsparse_types.h"
#include <torchsparse/sparse.h>
#include <lantern/types.h>

namespace make_raw {

// This is mostly the same as:
// return (void*) new tensor_pair(x); 
// but in a fancy C++ way.
void* TensorPair (const tensor_pair& x) {
  return make_ptr<tensor_pair>(x);
}
}

// This simply tells the compiler to consider that `void*` is a pointer to
// `tensor_pair` and then returns this reference.
namespace from_raw {
tensor_pair& TensorPair (void* x) {
  return *reinterpret_cast<tensor_pair*>(x);
}
}

// ---- there's more.
```

Additionally in this file we will implement functions that will allow us to free the memory pointed by this pointer, as well as tools to help us cast this type to something that we can return to R.

``` cpp
// Takes a void* pointer and deletes the memory it points to.
// First need to cast to the correct type.
// [[torch::export]]
void delete_tensor_pair(void* x) {
  delete reinterpret_cast<tensor_pair*>(x);
}

// Extract a single Tensor from this type. This will allow us to
// convert this type into a list of tensors to pass to R.

// [[torch::export]]
torch::Tensor tensor_pair_get_first(tensor_pair x) {
  return std::get<0>(x);
}

// [[torch::export]]
torch::Tensor tensor_pair_get_second(tensor_pair x) {
  return std::get<1>(x);
}
```

We are done with the type implementation on the `csrc` side. We can now implement a function that uses this type. For example:

``` cpp
// [[torch::export(register_types=c("tensor_pair", "TensorPair", "void*", "torchsparse::tensor_pair"))]]
tensor_pair sparse_relabel(torch::Tensor col, torch::Tensor idx) {
  return relabel(col, idx);
}
```

Note the arguments in the `// [[torch::export]]` special comment. Here we are telling `torchexport` what to generate. Namely, on encountering the type **`tensor_pair`** (first argument), it should:

-   Use `make_raw::TensorPair` and `from_raw::TensorPair` to cast between `void*` and `tensor_pair`. (`TensorPair` is the second argument.)
-   Use `void*` as the C type. In general it will always be `void*` here, unless you can cast to another C type and still re-create the object. (`void*` is the third argument.)
-   Use `torchsparse::tensor_pair` as the Rcpp type. (`torchsparse::tensor_pair` is the third argument.) This we haven't implemented yet; it'll be our next step.

**Note**: You only need to register the type once, for a single function that uses it. If we want to export other functions that return `tensor_pair` we won't need to register the type again in the `[[torch::export]]` comment.

We can now `cmake --build . --target install --parallel 8` to compile the `csrc` library, and everything should go fine.

### The `Rcpp` side

Now, on to the Rcpp side. We need to implement the custom Rcpp type that will hold the `void*` pointer returned from the `csrc` side, take care of the corresponding memory, and cast to `SEXP` when needed.

First we declare this type in `src/torchsparse_types.h`:

``` cpp
#pragma once
#include <torch.h>

namespace torchsparse {

class tensor_pair {
public:
  // this is the slot to hold the void*
  std::shared_ptr<void> ptr;
  // the constructor from a void*
  tensor_pair (void* x);
  // casting operator Rcpp->SEXP
  operator SEXP () const;
  // returns the void* from the type.
  void* get ();
};

}
```

The type having been declared, we can now implement its member functions in `src/torchsparse_types.cpp`:

``` cpp
#include <Rcpp.h>
#include "torchsparse_types.h"
#include "exports.h"

namespace torchsparse {

void* tensor_pair::get() {
  return ptr.get();
}

// Creates a list of two torch::Tensor's from this object.
tensor_pair::operator SEXP () const {
  Rcpp::List out;
  out.push_back(rcpp_tensor_pair_get_first(*this));
  out.push_back(rcpp_tensor_pair_get_second(*this));
  return out;
}

// initialize the `ptr` slot and **very important** register the custom
// deleter `rcpp_delete_tensor_pair` that will free the pointer's memory
// once `ptr` is gone (and hence once the `torchsparse::tensor_pair` instance
// is gone).
tensor_pair::tensor_pair (void* x) : ptr(x, rcpp_delete_tensor_pair) {};

}
```

You should now be able to `devtools::load_all()` and call `rcpp_sparse_relabel()`. This function will return a list with two `torch` Tensors.

In this case, we don't need to provide functionality for casting from a `SEXP` to the Rcpp type. If we had to, we'd have to implement the `tensor_pair::tensor_pair (SEXP x)` constructor, and probably have a function similar to `rcpp_tensor_pair_get_first`, but doing things the other way around: i.e., taking two `torch` Tensors and returning a `tensor_pair`.

Finally, this tutorial is more like a set of notes (from when I was implementing that functionality in `torchsparse`) than a "tutorial". Please open an issue if something is not clear, and we will be very happy to help!
